---
title: "Assignment 3: ACS via tidycensus"
author: "Avinash Singh"
format: html
execute:
  warning: false
  message: false
---

## Part 1: Set-up packages

### This chunk loads all the packages required for the assignment. It also tells tigris to reuse downloaded geographic boundaries.

```{r}

# Packages needed to retrieve, organize, map, and save ACS data
library(tidycensus) # Download ACS estimates and calculate margins of error
library(tigris)     # Download Census geographic boundaries
library(sf)         # Work with spatial data and map geometry
library(dplyr)      # Organize, calculate, select, and sort variables
library(tidyr)      # Reshape tidy data when necessary
library(ggplot2)    # Create the choropleth map
library(readr)      # Save the finished dataset as a CSV file

# Cache geographic boundaries so they do not need to be downloaded repeatedly
options(tigris_use_cache = TRUE)


```

## Part 2: Choose the geography and ACS variables

### This chunk searches the 2024 ACS five-year variable list and establishes the parameters for the analysis. The unit of analysis is the Texas county. Median household income will be used for the map, while the poverty count and its denominator will be used to calculate a county poverty rate.

```{r}

# Explore the variables available in the 2024 ACS 5-year dataset
# Do not use the deprecated `cache` argument
vars <- load_variables(2024, "acs5")

# Display the first 10 variables whose codes begin with B19
vars |>
    dplyr::filter(grepl("^B19", name)) |>
    dplyr::slice_head(n = 10)

# Select Texas as the state for the analysis
state_abbr <- "TX"

# Select counties as the geographic unit
geo_level <- "county" # options: state, county, tract, block group

# Select income, poverty count, and poverty denominator variables
my_vars <- c(
    income = "B19013_001", # median household income
    pov = "B17001_002",    # people below poverty
    povden = "B17001_001"  # poverty universe = denominator
)

# Select the required ACS year
year_acs <- 2024

# Select the ACS 5-year survey
survey <- "acs5"

```

```         
```

## Part 3: Download the data and calculate the poverty rate

### This chunk downloads the 2024 ACS estimates, margins of error, and geographic boundaries for Texas counties. The `E` suffix identifies an estimate, while the `M` suffix identifies its margin of error. It then calculates the poverty rate and propagates the uncertainty using `moe_prop()`.

```{r}

# Download the selected ACS variables and Texas county geometry
# Wide output produces one row per county with separate estimate and MOE columns
acs_wide <- get_acs(
    geography = geo_level,  # Use counties as the geographic unit
    variables = my_vars,    # Download the three selected ACS variables
    state = state_abbr,     # Limit the data to Texas
    year = year_acs,        # Use the 2024 ACS
    survey = survey,        # Use the ACS 5-year estimates
    geometry = TRUE,        # Retain county boundaries for mapping
    output = "wide"         # Produce one row for each county
)

# Derive the poverty rate and carry its margin of error through
acs_wide <- acs_wide |>
    mutate(
        pov_rate = povE / povdenE, # Divide poverty count by its denominator
        pov_moe = moe_prop(        # Calculate the MOE for the poverty rate
            povE,                  # Poverty-count estimate
            povdenE,               # Poverty-denominator estimate
            povM,                  # MOE for the poverty-count estimate
            povdenM                # MOE for the denominator estimate
        )
    )

# Check the completed spatial dataset
acs_wide
nrow(acs_wide)
names(acs_wide)
```

## Part 4: Produce the map, tables, uncertainty check, and CSV file

### This chunk produces all the main assignment outputs:

-   A choropleth map of median household income

-   The ten counties with the highest poverty rates

-   The ten counties with the lowest poverty rates

-   Poverty-rate margins of error

-   Confidence-interval limits for evaluating the rankings

-   A CSV file containing the completed ACS dataset

```{r}

# Create a choropleth map of median household income
ggplot(acs_wide) +
    geom_sf(
        aes(fill = incomeE), # Fill each county according to its income estimate
        color = NA           # Remove county-border lines
    ) +
    scale_fill_viridis_c(
        name = "Median HH Income", # Give the map legend a clear title
        labels = scales::label_dollar()
    ) +
    labs(
        title = paste0(
            "ACS ", year_acs,
            " 5-year: Median Income — ",
            state_abbr, " (", geo_level, ")"
        ),
        caption = "Source: U.S. Census Bureau via tidycensus"
    ) +
    theme_minimal() +
    theme(
        text = element_text(family = "Palatino"), # Drop family if unavailable
        plot.title = element_text(size = 14, face = "bold"),
        plot.caption = element_text(size = 8),
        legend.position = "inside",
        legend.position.inside = c(0.2, 0.15)
    )


# Create a table of the 10 counties with the highest poverty rates
top10 <- acs_wide |>
    st_drop_geometry() |>            # Remove geometry because it is not needed
    arrange(desc(pov_rate)) |>       # Sort from highest to lowest poverty rate
    select(
        NAME,
        pov_rate,
        pov_moe,
        povE,
        povM
    ) |>
    slice_head(n = 10)               # Keep the first 10 counties


# Create a table of the 10 counties with the lowest poverty rates
bottom10 <- acs_wide |>
    st_drop_geometry() |>            # Remove geometry because it is not needed
    arrange(pov_rate) |>             # Sort from lowest to highest poverty rate
    select(
        NAME,
        pov_rate,
        pov_moe,
        povE,
        povM
    ) |>
    slice_head(n = 10)               # Keep the first 10 counties


# Display the two tables
top10
bottom10


# Add confidence-interval limits to help check uncertainty in the rankings
ranked_areas <- bind_rows(
    top10 |> mutate(table_group = "Highest poverty rates"),
    bottom10 |> mutate(table_group = "Lowest poverty rates")
) |>
    mutate(
        pov_lower = pmax(0, pov_rate - pov_moe), # Lower confidence limit
        pov_upper = pmin(1, pov_rate + pov_moe)  # Upper confidence limit
    ) |>
    select(
        table_group,
        NAME,
        pov_rate,
        pov_moe,
        pov_lower,
        pov_upper
    )

# Display rates, MOEs, and confidence-interval limits
ranked_areas


# Save the completed nonspatial data as a CSV file
write_csv(
    st_drop_geometry(acs_wide),
    paste0("acs_", state_abbr, "_", year_acs, ".csv")
)
```

## 4(a) Map Interpretation

The map shows substantial variation in median household income across Texas counties. Higher-income counties appear primarily around major metropolitan areas, especially the Dallas–Fort Worth, Austin, and Houston regions. Many counties in South Texas, along the Rio Grande border, and in parts of West Texas have lower median household incomes. Overall, the pattern suggests that counties near large urban economies tend to have higher household incomes than many rural and border counties.

## 4(b) Table Interpretation

The highest estimated poverty rate is in Dimmit County at approximately 41.1%, followed by Zapata County at 38.4% and Zavala County at 33.9%. Many counties in the highest-poverty group are located in South Texas or near the Texas–Mexico border. In contrast, Rockwall County has the lowest estimated poverty rate at approximately 4.0%, followed by Sterling County at 4.3% and Sutton County at 4.5%. The margins of error vary substantially across counties, meaning the estimated ordering should be interpreted cautiously, particularly for counties with relatively large margins of error.\

## 4(c) Uncertainty Check

Dimmit County and Zapata County provide one clear example of overlapping confidence intervals.

-   Dimmit County: estimated poverty rate = 41.1%, MOE = 8.9 percentage points, interval = 32.2%–50.0%.

-   Zapata County: estimated poverty rate = 38.4%, MOE = 6.0 percentage points, interval = 32.4%–44.4%.

The two intervals overlap substantially. Therefore, although Dimmit County is ranked above Zapata County based on the point estimates, the data do not provide strong evidence that Dimmit County’s true poverty rate is higher. Sampling uncertainty could reverse or eliminate the apparent difference. The ranking should consequently be treated as descriptive rather than as a definitive ordering of the counties.

## 4(d) Methods Note

This analysis uses 2024 American Community Survey five-year estimates for Texas counties. Counties were selected because they provide meaningful geographic comparisons while remaining readable on a statewide map. Median household income (`B19013_001`) was mapped to show the geographic distribution of economic resources. Poverty was measured using the population below the poverty level (`B17001_002`) divided by the population for whom poverty status was determined (`B17001_001`). Using a rate instead of a raw count prevents counties with larger populations from appearing more disadvantaged simply because they contain more residents. I retained the ACS margins of error and used `moe_prop()` to propagate uncertainty into the calculated poverty rate. The margins of error show that some county estimates are much less precise than others. Several confidence intervals overlap, so small differences between counties should not be interpreted as definitive differences in their true poverty rates. Consequently, the top-and-bottom rankings are descriptive and should be interpreted cautiously.
