---
title: "Causal Structure of Organizational Lethality"
subtitle: "Based on Asal & Rethemeyer (2008), *The Nature of the Beast*"
author: "Avinash"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

## Setup

```{r}
library(ggdag)
library(dagitty)
library(ggplot2)
```

## What is causal inference?

Causal inference asks not "are X and Y correlated" but "if I intervened and changed X, would Y change as a result" — it's the toolkit for answering that under real-world data limitations, not just experiments.

## Research question

What organizational characteristics *cause* a terrorist organization to be more lethal? (Asal & Rethemeyer 2008)

## Build the DAG

```{r}
dag <- dagify(
  L  ~ I + Sz + SS + Ag + C + T + R + W,   # Lethality caused by all these
  T  ~ R,                                   # Regime type -> territorial control
  C  ~ Sz,                                  # Size -> alliance connections
  D  ~ I + Sz,                              # Selection into dataset (the collider)
  exposure = "T",
  outcome  = "L",
  labels = c(
    I  = "Ideology",
    Sz = "Org Size",
    SS = "State\nSponsorship",
    Ag = "Org Age",
    C  = "Alliance\nConnections",
    T  = "Territorial\nControl",
    R  = "Regime\nType",
    W  = "Wealth",
    D  = "Dataset\nSelection",
    L  = "Lethality"
  )
)
```

## Basic diagram

```{r}
ggdag(dag, text = FALSE, use_labels = "label") +
  theme_dag() +
  ggtitle("Asal & Rethemeyer (2008): Organizational Lethality DAG")
```

## Confounder check: paths between Territorial Control and Lethality

Regime Type (R) is a common cause of both Territorial Control (T) and Lethality (L) directly, creating a backdoor path T \<- R -\> L.

```{r}
ggdag_paths(dag, from = "T", to = "L", text = FALSE, use_labels = "label") +
  theme_dag() +
  ggtitle("Open paths between Territorial Control and Lethality")
```

## Required adjustment set

This computes, from the DAG structure, which variables must be controlled for to identify the causal effect of Territorial Control on Lethality.

```{r}
ggdag_adjustment_set(dag, exposure = "T", outcome = "L",
                      text = FALSE, use_labels = "label") +
  theme_dag() +
  ggtitle("Adjustment set to identify effect of Territorial Control on Lethality")
```

## Collider check

`D` (whether an organization appears in the TKB dataset at all) is caused by both Ideology and Size — a collider created by sample selection, since TKB only contains organizations that already committed a violent act.

```{r}
ggdag_collider(dag, text = FALSE, use_labels = "label") +
  theme_dag() +
  ggtitle("Collider check")
```

## Notes

-   **Confounder (Regime Type):** must be adjusted for — it is a common cause of Territorial Control and Lethality.
-   **Collider (Dataset Selection):** must NOT be adjusted for. Because TKB only includes organizations that already used violence, the entire study is structurally conditioned on this collider — a limitation Asal & Rethemeyer acknowledge directly, noting that unclaimed/nonviolent religious organizations are systematically absent from the sample.
